Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-05-23 manuscript/Chapter2-manuscript.Rmd
update intro based on PO and Julien comments. update refs

---
title: "Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame"
output: 
  word_document:
    reference_docx: Chapter2-styles.docx
    fig_caption: yes
bibliography: references2.bib
csl: ecology.csl
---

@@

Many predator-prey systems are driven by reciprocal interactions where one responds to changes in the behaviour of the other. However, the mechanisms driving these individual responses are only beginning to be explored, in part because of the challenges associated with monitoring direct predator-prey interactions. In this study, we used a virtual predator-prey system (the online game *Dead by Daylight*) to test how hunting experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. With increasing experience, predators fine-tuned their tactics to the behaviour of their prey. Individuals that adopted a flexible tactic (i.e. generalists) encountered groups of prey with varying speeds, and those that adopted a cursorial and specialized tactic (i.e. specialists) encountered groups of prey with similar speeds throughout the study period. Our results suggest that experience enables predators to learn how to optimize their tactics to the type of prey that they encounter (i.e. fast or slow), which may act as a stabilizing mechanism promoting behavioural diversification in predator-prey systems.

@@

Individual variation in predator foraging behaviour is increasingly recognized as a major driver of trophic interactions and community dynamics [@Griffen.etal2012;@Michalko.Pekar2016;@Moran.etal2017;@Michalko.etal2021]. Indeed, predator populations often comprise assemblages of individuals specializing in different foraging tactics, with consequences for habitat use, functional responses, prey choice, and foraging rate [@Kobler.etal2009;@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@Matsumura.Miyatake2022a]. Despite the clear evidence that individual specialization is important for predator-prey interactions, its underlying mechanisms remain surprisingly unclear. However, it has recently been suggested that reciprocal interactions between predators and prey could drive individual foraging specialization [@Araujo.etal2011;@toscanoPersonalityForagingBehavior2016;@DeLisle.etal2022]. For instance, predators often hunt using techniques that are fine-tuned to the type of prey that they encounter [@Estes.etal2003;@Woo.etal2008;@Courbin.etal2018], and their refinement is crucial for young predators to reach adulthood and survive [@Phillips.etal2017;@Heithaus.etal2018]. Thus, the capacity of predators to effectively use hunting tactics should be contingent on periods of extensive practice (i.e. experience), which may result in different foraging specializations.

Theory predicts that foraging specialization may emerge via learning and expertise acquisition [@Tinker.etal2009;@Dukas2019]. @Dukas2019 defines expertise as the characteristics, skills, and knowledge allowing individuals with extensive experience to outperform novices on complex tasks. It is described as the relationship between success and experience, where extensive practice optimizes success when individuals reach an advanced level of experience. Empirical studies on human and non-human hunters show that experience optimizes foraging efficiency (e.g. search and handling times, return rates) potentially via associative images or reliance on prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. This body of work suggests that such optimization may thus promote individual foraging specialization (e.g. repeatedly using fast attacks) if the success of a tactic is constant through time. It may also be costly to test different tactics by trial and error when prey are scarce or unpredictable. Alternatively, the costs/risks faced by hunters when switching tactics may be offset through extensive practice, as individuals gain experience and information on their prey, leading to increased individual foraging flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Kelley.Magurran2011;@Snell-Rood2013]. To develop proper responses to fluctuating resources, individuals would need to sample their environment broadly during an extensive period of time, leading to greater performance at later stages of development [reviewed in @Snell-Rood2013]. However, predators can differ in the challenges that they face during their lifetime, which could result in both processes (i.e. specialization and flexibility) operating at the same time within a predator population. For instance, some individuals may hunt in habitats where prey are faster, which are often proven to be more difficult to hunt [@Walker.etal2005;@Kelley.Magurran2011;@Martin.etal2022]. It should thus be essential for predators to learn how to match their tactic to their prey to optimize their success.

There is currently a lack of consensus on the fitness advantages of specialized vs flexible foraging [@Phillips.etal2017]. This is reflected in the literature showing contrasting results in the links between specialization and fitness. Some studies report greater benefits when specializing [@Pintor.etal2014;@Patrick.Weimerskirch2014a;@vandenBosch.etal2019], some report that flexible foraging has greater benefits [@Paull.etal2012;@Santoro.etal2019;@Manlick.etal2021], and others find equal benefits depending on timescales [@Woo.etal2008;@Potier.etal2015]. This may in part be due to the limited information we have on the ecological contexts favoring specialization over flexibility. However, in predator-prey systems, studies suggest that fluctuations in the predictability of resources throughout a predator's lifetime may be a key factor shaping foraging specialization [@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017;@Courbin.etal2018]. It is hypothesized that when resources are predictable (or stable), individual specialists should have higher capture rates by reducing the energy and time required to search for and handle prey. In contrast, individual generalists should benefit from adjusting to fluctuating resources by reducing the consequences of uncertainty. While it is clear that predators can match their behaviour and locomotion to their prey [@McGhee.etal2013a;@Bro-Jorgensen2013;@Chang.etal2017;@Szopa-Comley.Ioannou2022], the important question remaining is how? Given the dynamic nature of predator-prey interactions, it is unlikely that evolution alone could explain such behavioural adjustments by predators, because complex hunting strategies require practice to be effective [@Kelley.Magurran2011]. Thus, uncovering the role of learning during direct predator-prey interactions would help in better predicting the behavioural decisions and the success of predators when they are hunting.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and prey. We recently demonstrated, using an online predator-prey videogame called *Dead by Daylight* (*DBD*), that virtual systems can overcome this challenge [@FraserFranco.etal2022]. In this dynamic virtual system, four prey players need to forage for resources while avoiding a predator player. The prey can contribute to the group's success by healing or helping others escape the predator, or be selfish and attempt to escape by themselves [@Cere.etal2021]. The predators' hunting tactics and success are driven by the movement of the prey [@FraserFranco.etal2022]. Predators should thus benefit from learning how the prey behave to be successful. Other virtual predator-prey studies show that predation regimes drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al. in prep). These examples outline how virtual systems could be useful in identifying general ecological patterns, which might in turn help to advance the current gaps in predator-prey research. They also generate large volumes of data on interacting players throughout their lifetime in the game under realistic ecological scenarios. Hence, virtual systems offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we test how hunting experience shapes predator foraging specialization using data from players in the online videogame *Dead by Daylight*. *DBD* simulates a direct predator-prey interaction, where one predator player hunts four prey players in different virtual environments. The data grants a high degree of precision on the behavioural interaction, as the behaviour of both the predator and the four prey along with the predator's success are monitored simultaneously in each trial. First, we investigate how predators develop their individual hunting expertise. We hypothesize that predators should differ in the development of their expertise, partly because they encounter varying levels of difficulty with the prey that they pursue. For example, a predator may face greater difficulty than other individuals if it more often encountered groups of prey that were elusive. Second, we test the hypothesis that experience will shape foraging specialization. If experience reduces the costs of switching between hunting tactics, we predict that the predator population should become more flexible. Alternatively, if experience enables the refinement of the hunting tactics, then the predator population may instead specialize. Otherwise, if both processes operate at the same time, then we should see no change with experience at the population level, but differences in specialization among individuals should increase. Third, we evaluate how predator foraging specialization emerges from behavioural interactions with the prey. We expect predators that experienced more predictable encounters with their prey to specialize, while predators that experienced unpredictable encounters with their prey should adopt a flexible hunting tactic. If we detect such prey-dependent fine-tuning, then specialist and flexible hunters should attain equal success because both should have learned to optimize their hunting style to their prey.

>>>>

---
title: "Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame"
output: 
  word_document:
    reference_docx: Chapter2-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: ecology.csl
---

@@

Many predator-prey systems are driven by reciprocal interactions where one responds to changes in the behaviour of the other. However, the mechanisms driving these individual responses are only beginning to be explored, in part because of the challenges associated with monitoring direct predator-prey interactions. In this study, we used a virtual predator-prey system (the online game *Dead by Daylight*) to investigate if and how hunting experience shapes individual foraging specialization and success in predators across repeated interactions with their prey.

*à modifier pour parler plutôt de l'interaction qui devient fine tuned (i.e. predators and prey)*
We found that with increasing experience, predators fine-tuned their tactics to the behaviour of their prey. Individuals that adopted a flexible tactic (i.e. generalists) encountered groups of prey with varying speeds, and those that adopted a cursorial and specialized tactic (i.e. specialists) encountered groups of prey with similar speeds throughout the study period.

Our results suggest that experience enables predators to learn how to optimize their tactics to the type of prey that they encounter (i.e. fast or slow), which may act as a stabilizing mechanism promoting behavioural diversification in predator-prey systems.

@@

Individual variation in predator foraging behaviour is increasingly recognized as a major driver of trophic interactions and community dynamics [@Griffen.etal2012;@Michalko.Pekar2016;@Moran.etal2017;@Michalko.etal2021]. Indeed, predator populations often comprise assemblages of individuals specializing on different foraging tactics, with consequences for habitat use, functional responses, prey choice, and foraging rate [@Kobler.etal2009;@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@Matsumura.Miyatake2022a]. Despite the clear evidence that individual specialization is important for predator-prey interactions, its underlying mechanisms remain surprisingly unclear. It has recently been suggested, however, that reciprocal interactions between predators and prey could drive individual foraging specialization [@Araujo.etal2011;@toscanoPersonalityForagingBehavior2016;@DeLisle.etal2022]. For instance, predators often hunt using techniques that are fine-tuned to the type of prey that they encounter [@Estes.etal2003;@Woo.etal2008;@Courbin.etal2018], and their refinement is crucial for young predators to reach adulthood and survive [@Phillips.etal2017;@Heithaus.etal2018]. Thus, the capacity of predators to effectively use hunting tactics should be contingent on periods of extensive practice (i.e. experience), which may result in different foraging specializations.

Theory predicts that foraging specialization may emerge via learning and expertise acquisition [@Tinker.etal2009;@Dukas2019]. @Dukas2019 defines expertise as the characteristics, skills, and knowledge allowing individuals with extensive experience to outperform novices on complex tasks. It is described as the relationship between success and experience, where extensive practice optimizes success when individuals reach an advanced level of experience. Empirical studies on human and non-human hunters show that experience optimizes foraging efficiency (e.g. search and handling times, return rates) potentially via associative images or reliance on prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. This body of work suggests that such optimization may thus promote individual foraging specialization (e.g. repeatedly using fast attacks) if the success of a tactic is constant through time. It may also be costly to test different tactics by trial and error when prey are scarce or unpredictable. Alternatively, the costs/risks faced by hunters when switching tactics may be offset through extensive practice, as individuals gain experience and information on their prey, leading to increased individual foraging flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Kelley.Magurran2011;@Snell-Rood2013]. To develop proper responses to fluctuating resources, individuals would need to sample their environment broadly during an extensive period of time, leading to greater performance at later stages of development [reviewed in @Snell-Rood2013]. However, predators can differ in the challenges that they face during their lifetime, which could result in both processes (i.e. specialization and flexibility) operating at the same time within a predator population. For instance, some individuals may hunt in habitats where prey are faster, which are often proven to be more difficult to hunt [@Walker.etal2005;@Kelley.Magurran2011;@Martin.etal2022]. It should thus be essential for predators to learn how to adjust their tactic to their prey to optimize their success.

There is currently a lack of consensus on the fitness advantages of specialized vs flexible foraging strategies [@Phillips.etal2017]. This is reflected in the literature showing contrasting results in the links between specialization and fitness at the individual level. Some studies have found greater success for individuals that use specialized foraging tactics [@Pintor.etal2014;@Patrick.Weimerskirch2014a;@vandenBosch.etal2019] while others found that individuals using flexible foraging tactics have greater success [@Paull.etal2012;@Santoro.etal2019;@Manlick.etal2021]. This disrepancy may in part be due to the limited information we have on the ecological contexts favoring specialization over flexibility. However, many studies have found that both type of foragers can also attain equal success depending on timescales, and suggest that fluctuating resources throughout a predator's lifetime may be key in shaping individual foraging specialization [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015;@Phillips.etal2017;@Courbin.etal2018]. It is hypothesized that when resources are predictable (or stable), individual specialists should have higher capture rates by reducing the energy and time required to search for and handle prey. In contrast, individual generalists should benefit from adjusting to fluctuating resources by reducing the consequences of uncertainty [@dall2010]. While it is clear that predators can match their hunting behaviour and locomotion to their prey [@McGhee.etal2013a;@Bro-Jorgensen2013;@Chang.etal2017;@Szopa-Comley.Ioannou2022], the important question remaining is how? Given the dynamic nature of predator-prey interactions, it is unlikely that evolution alone could explain such behavioural adjustments by predators, because complex hunting strategies require practice to be effective [@Kelley.Magurran2011]. Thus, uncovering the role of learning during direct predator-prey interactions would help in better predicting the behavioural decisions and the success of predators when they are hunting.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and prey. We recently showed, using an online predator-prey videogame called *Dead by Daylight* (*DBD*), that virtual systems can overcome this challenge [@FraserFranco.etal2022]. In this dynamic virtual system, four prey players need to forage for resources while avoiding predation by a predator player. The prey can contribute to the group's success by healing or helping others escape the predator, or be selfish and attempt to escape by themselves [@Cere.etal2021]. The predators' hunting tactics and success are driven by the movement of the prey [@FraserFranco.etal2022]. Predators should thus benefit from learning how the prey behave to be successful. Other virtual predator-prey studies show that predation regimes drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al. in prep). These examples outline how virtual systems could be useful in identifying general ecological patterns, which might in turn help to advance predator-prey research. They also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems, including *DBD*, offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we evaluated how hunting experience shapes predator foraging specialization and success using data from players in the online videogame *Dead by Daylight*. First, we investigate how predators develop their individual hunting expertise. We hypothesize that predators should differ in the development of their expertise if they encountered varying levels of difficulty with the prey that they pursued. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we test the hypothesis that experience will shape individual foraging specialization. We define foraging tactics as the mean movement speed of the predator along a slow-fast continuum, and the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed from one match to the next where low IIV translates to individual specialization. If experience reduces the costs of switching between low and high speeds, we predict that individuals (and thus the population) should all converge towards flexible speeds. If experience enables the refinement of the tactics to hunt specific types of prey, then all individuals may instead specialize in their speed. In both scenarios, among individual variation in IIV (i.e. individual specialization) should be low because everyone is either a specialist or a generalist, but the population variance should either increase (flexibility) or decrease (specialization). Alternatively, both processes may operate at the same time if behavioural interactions with the prey vary among predators, such that among individual variation in IIV should increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. homogeneous groups across matches) to specialize in their speed, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then specialist and flexible hunters should attain equal success because both should have learned to optimize their hunting style to their prey.
